Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy

### 2023-11-04 manuscript/manuscript.qmd
More Suparna edits

Over the last three decades, governments globally have increasingly sought to crack down on NGOs (CITE). Repressive governments have repressed these organizations using both violence, as well as administrative crackdown, which create barriers to funding and advocacy for a variety of NGOs. (CITE) Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" (CITE) and data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in nearly 60% of countries---111 in total (UPDATE FROM NEW CIVICUS DATA). These attacks---most of which have been implemented through anti-NGO laws---have adversely affected the ability of both domestic and international NGOs to operate and raise funds. Official aid funds have decreased substantially in countries that have imposed restrictive anti-NGO legislation [@ChaudhryHeiss:2018; @DupuyPrakash:2018] and official aid donors channel fewer funds to contentious issues and instead choose to fund "tame" causes [@ChaudhryHeiss:2018; @Bush:2015; @Herrold:2020; @RightSpringmanWibbels:2022]. In response to this dramatic shift in the NGO funding landscape, international NGOs (INGOs) have increasingly turned to foreign individual donors to offset the loss of funding from government and foundation sources [@BanksHulmeEdwards:2015].

@@

Although a robust body of work on NGOs and donor behavior examines the motivation of public donors, private foundations and elite donors (CITE), we are yet to examine the impact of closing civic space, especially individual private donors, who are likely to be important funders in the face of increased government repression. Existing research shows that organizational accountability and financial transparency play an important role in motivating individual donors to donate to NGOs (CITE). In this article, we extend this literature by evaluating how these organization-level characteristics influence private donor behavior in an era of shrinking civic space. In considering this question, we engage with a growing body of research on the consequences of closing civic space on NGOs and donors (CITE). We argue that NGO-level organizational factors and domestic political environments of NGO host countries serve as heuristics and can influence foreign private donors' preferences. We fielded a conjoint experiment to a nationally representative sample of X adults in the US. A conjoint design allows us to directly compare the relative impact of host government repression of NGOs vis-a-vis an organization's financial transparency and accountability on individual donor preferences. The result of these experiments show both financial transparency and accountability increases individuals' willingness to donate NGOs. Learning about host government criticism and crackdown against NGOs decreases the likelihood of donation by itself; however, organizational features such as financial transparency and accountability protect against this dampening effect. INSERT LINE ABOUT EFFECT SIZE.

@@

Substantial research on NGOs has looked at the motivation of individual donors, including altruism, reputational benefits, and alignment with personal values [@BilodeauSlivinski:1997; @Wiepking:2010; @BekkersWiepking:2011]. However, this research has overwhelmingly looked at individual donor behavior and motivation for giving to local---rather than international---causes (CITE -2). Understanding the latter is important as philanthropy to international causes can be harder to motivate because the number of recipients is larger and further removed from the donor [@CasaleBaumann:2015, 100]. Further, as @MicklewrightSchnepf:2009 point out, motivations for donating to ongoing international causes differ from donating money to natural disasters or emergency international causes. Under ordinary circumstances, "donors are more likely to donate to a charity operating locally than to a charity providing identical service abroad" [@Tremblay-BoirePrakash:2017, 644]. However, we lack adequate data and theory regarding philanthropy towards INGOs.

The third wave of democratization, the collapse of the Soviet Union, and the spread of a "pro-NGO norm" led to an increased role and visibility for NGOs. Globally, a large amount of resources were channeled to NGOs focused on service provision or development because many public and private donors perceived them as more efficient than states with weak institutions and poor governance; there was a high risk that such states would use aid transfers inefficiently [@Dietrich:2013].

@@

Donors respond to a host of heuristics and signals when deciding whether to donate to an NGO, since seeking out complete information about an organization's deservingness and efficiency is costly and time-consuming [@CrosonShang:2011; @Tremblay-BoirePrakash:2017]. Nonprofits fundraising internationally may therefore rely on framing to control perceptions. Below, we theorize the effect of different frames surrounding a country's legal environment and nonprofit organizational characteristics on individual donor preferences.

TODO: Cite @ChaudhryHeiss:2021 somewhere

@@

One common signal for donors is organizational efficacy or the belief that an individual's donation will make a real difference. This can manifest itself through multiple mechanisms such as the amount an organization spends on overhead costs, whether an organization makes its finances public, or whether an organization participates in voluntary accountability practices, such as listing organizational details on websites like GuideStar and Charity Navigator. An NGO's financial transparency, defined as the degree of completeness of information, provided by organizations to the public concerning its business activities, is one tangible heuristics that donors use to make their decisions. NGOs can engage in financial transparency in a variety of ways such as distribution of audited financial statements, third party intermediaries. Research shows that transparency is associated with greater donations, lower debt and better governance [@HarrisNeely:2021; @SaxtonKuoHo:2012]. It can also enhance donor trust and confidence in organizations [@Vaccaro:2006; @VaccaroMadsen:2006].

@@

In addition to organizational factors discussed above, political contexts in NGO host countries, especially their environments for civil society organizations, can also influence donor preferences for donating to foreign NGOs. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations [@GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022; @Heiss:2017; @DupuyRonPrakash:2015]. As a result, NGOs must spend more time, effort, and resources to ensure their survival—at the expense of pursuing their missions. Some NGOs have adapted to their repressive environments by recreating their organizations and changing their mission to avoid directly confronting the government. Analyzing the impact of anti-NGO laws in Bangladesh and Zambia, scholars have similarly found that NGOs with a broad mission ended up restricting themselves to service work and giving up on their advocacy work. Those that continued to engage in transnational advocacy changed the targets of their advocacy, the issues they focused on, and the language they used. For instance, many NGOs stopped lobbying the central government and focused their advocacy efforts on lower levels of government [@FransenDupuyHinfelaar:2021]. Similarly, in Russia, many NGOs working with foreign partners and lobbying the central government no longer consider it an effective strategy [@SundstromHenrySperling:2022]. Overall, fewer groups may end up focusing on contentious causes, instead focusing on regime-compatible causes.

The changes in civic environments impacting NGO activity have also influenced official donor responses. In response to the passage of these laws, scholars have found that while multilateral donors do not reduce aid following the passage of anti-NGO laws, they are associated with a 32% decline in bilateral aid inflows [@DupuyPrakash:2018]. Further, advocacy-oriented donors---i.e., those that fund democracy and civil society promotion activities---reduced their spending by 74%, but did not cut their spending on development projects such as education, health, water, and sanitation [@RightSpringmanWibbels:2022]. However, previous literature has established that individual private donors may not necessarily use the same metrics as official donors (CITE). Private foreign aid does not have to deal with the same strategic and political considerations as official aid and therefore, may be in a better position to respond to recipients' needs on the grounds [@EasterlyWilliamson:2011]. These anti-NGO laws may signal to private donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and crack down on them. Therefore, we expect that government restrictions abroad will increase respondents' likelihood of donating to the organization.

@@

Identifying the causal link between structural factors and donor behavior is more difficult than measuring the link between organizational factors and donor behavior, however, because (1) NGOs have less direct control over their political environments, and (2) organizational characteristics like financial transparency and accountability can sour the NGO-government relationship---governments may be more likely to target NGOs that lack financial transparency and accountability. Legal restrictions do not occur in isolation. Countries are more likely to repress NGOs when organizational features are in tension with government preferences and when organizations receive substantial funding from abroad [@DupuyRonPrakash:2015]. Therefore, the same organizational characteristic that can increase the likelihood of crackdown can also simultaneously serve as a donor heuristic and influence perceptions of NGO deservingness.

@@

A conjoint research design also allows us to calculate treatment effects even if respondents do not see every combination of organization attributes, and correspondingly provides a substantial increase in statistical power. In standard factorial experimental designs, every combination of treatment attributes would need to be shown to some number of participants would ordinarily need to be shown all possible combinations of experimental attributes. In our experiment, we presented respondents with six possible treatments with varying counts of attributes for each treatment, yielding 576 possible unique combinations of features. Even with respondents answering twelve iterations of the experiment, not every combination was seen. However, as long as all the possible organizational attributes are well randomized and there are no systematic biases toward specific options (i.e. more respondents select the first option because it is the first) or toward earlier iterations of attribute choices (i.e. respondents are more careful and attentive for the first hypothetical organization than the last), we can pool all observations together for specific attributes of interest while marginalizing across all other attributes [@KertzerRenshonYarhi-Milo:2021]. This allows us to (1) estimate the effect of each experimental treatment even if some unique combinations of attributes were unseen, and (2) use a much smaller sample size than would be required in a more traditional factorial design.

@@

# Results and discussion

@@

## Effect of general things (lol needs better heading)

@@

Over the last two decades, governments globally have systematically controlled and repressed NGOs using both violent tactics and administrative means over the last three decades---a phenomenon known as "closing" or "shrinking" civic space. This trend of closing civic space has important implications for local and international NGOs as well as donors. Recent research has analyzed the impact of this crackdown on public or official aid donors. However, we know relatively little about how state repression of NGOs affects the preferences of foreign private donors. The responses of elite donors and foundations to government crackdown on civil society organizations remains a ripe area for research. Unlike state donors, individual and elite donors may not necessarily withdraw or reduce support for NGOs facing harassment abroad. This is important because private philanthropy towards organizations working in international affairs continues to grow despite inflation---from 2021 to 2022, individual giving to nonprofits in international affairs grew by over 10.9% in the US [@GivingUSA:2023].

Given the increasing hurdles faced by these nonprofits, how do individual donors in the U.S. feel about donating to legally besieged NGOs abroad? How do organizational-level factors such as financial transparency and accountability compare with more structural-level factors such as host country politics when deciding to donate? Using a conjoint experiment with a nationally representative sample of the U.S. population, we find COMPLETE.

These results have important implications for NGO operations, fundraising and survivability at a time when a number of INGOs are dealing with hostile host governments. The crackdown is even more challenging for local NGOs in Global South countries, where much government and public support is channeled via service delivery organizations. Local groups tackling contentious issues may not be able to raise funds domestically, as publics may prefer donating to service organizations focusing on health, education, and poverty [@AbsarCrowRon:2017]. Even local philanthropists may be deterred from giving to organizations focusing on contentious causes such as advocacy, media freedom, and anti-corruption initiatives due to poor tax incentives or a fear of retribution [@Brechenmacher:2017; @DupuyRonPrakash:2015; @Baoumi:2016]. Our results highlight the importance of organizational characteristics like transparency and accountability even in an era of closing civic space. By finding new ways to communicate about organizational transparency and accountability to potential donors, NGOs can help overcome individual donors' concerns about host governments' criticisms of and crackdown on these groups.

Future research should test how well these results map on to elite individual donors, not just the average private donor, as well as determine the extent to which these preferences translate into behavior.

>>>>

Over the last three decades, governments globally have increasingly sought to crack down on NGOs [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016, @Heiss:2017; @GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022]. Repressive governments have cracked down on these organizations using both violence, as well as administrative crackdown, which create barriers to funding and advocacy for a variety of NGOs [@DupuyRonPrakash:2016; Chaudhry 2022]. Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" [@CarothersBrechenmacher:2014; @Rutzen:2015] and data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in more than 60% of countries---117 in total (https://monitor.civicus.org/globalfindings/). These attacks---most of which have been implemented through anti-NGO laws---have adversely affected the ability of both domestic and international NGOs to operate and raise funds. Official aid funds have decreased substantially in countries that have imposed restrictive anti-NGO legislation [@ChaudhryHeiss:2018; @DupuyPrakash:2018] and official aid donors channel fewer funds to contentious issues and instead choose to fund "tame" causes [@ChaudhryHeiss:2018; @Bush:2015; @Herrold:2020; @RightSpringmanWibbels:2022]. In response to this dramatic shift in the NGO funding landscape, international NGOs (INGOs) have increasingly turned to foreign individual donors to offset the loss of funding from government and foundation sources [@BanksHulmeEdwards:2015].

@@

Existing research shows that organizational accountability and financial transparency play an important role in motivating individual donors to donate to NGOs [@HarrisNeely:2021; @SaxtonKuoHo:2012; @Vaccaro:2006; @VaccaroMadsen:2006; @Tremblay-BoirePrakash:2017; @Becker:2018]. In this article, we extend this literature by evaluating how these organization-level characteristics influence private donor behavior in an era of shrinking civic space. In considering this question, we engage with a growing body of research on the consequences of closing civic space on NGOs and donors [@DupuyPrakash:2018; @RightSpringmanWibbels:2022; @FransenDupuyHinfelaar:2021; @SundstromHenrySperling:2022; @Herrold:2020; @ChaudhryHeiss:2021]. We argue that NGO-level organizational factors and domestic political environments of NGO host countries serve as heuristics and can influence foreign private donors' preferences. We fielded a conjoint experiment to a nationally representative sample of X [HOW MANY] adults in the US. A conjoint design allows us to directly compare the relative impact of host government repression of NGOs vis-a-vis an organization's financial transparency and accountability on individual donor preferences. The result of these experiments show both financial transparency and accountability increases individuals' willingness to donate NGOs. Learning about host government criticism and crackdown against NGOs decreases the likelihood of donation by itself; however, organizational features such as financial transparency and accountability protect against this dampening effect. INSERT LINE ABOUT EFFECT SIZE.

@@

Substantial research on NGOs has looked at the motivation of individual donors, including altruism, reputational benefits, and alignment with personal values [@BilodeauSlivinski:1997; @Wiepking:2010; @BekkersWiepking:2011]. However, this research has overwhelmingly looked at individual donor behavior and motivation for giving to local---rather than international---causes [@BekkersWiepking:2011; @Tremblay-BoirePrakash:2017]. Understanding the latter is important as philanthropy to international causes can be harder to motivate because the number of recipients is larger and further removed from the donor [@CasaleBaumann:2015, 100]. Further, as @MicklewrightSchnepf:2009 point out, motivations for donating to ongoing international causes differ from donating money to natural disasters or emergency international causes. Under ordinary circumstances, "donors are more likely to donate to a charity operating locally than to a charity providing identical service abroad" [@Tremblay-BoirePrakash:2017, 644]. However, we lack adequate data and theory regarding the conditions under which private donors engage in increased philanthropy towards INGOs. This gap also exists because conventionally, public donors or official aid donors have been among the main funders of INGOs.

The third wave of democratization, the collapse of the Soviet Union, and the spread of a "pro-NGO norm" led to an increased role and visibility for NGOs. Globally, a large amount of official aid resources were channeled to NGOs focusing on service provision and development. public and private donors perceived them as more efficient than states with weak institutions and poor governance; there was a high risk that such states would use aid transfers inefficiently [@Dietrich:2013]. Less bureaucratic and more nimble than state institutions, NGOs were seen as key agents of development and democratization [@Herrold:2020, 142--43].

@@

Private donors, however, do not necessarily seek out complete information about an organization's deservingness and efficiency, since this is a costly and time-consuming process [@CrosonShang:2011; @Tremblay-BoirePrakash:2017]. As such, these donors respond to a host of heuristics and signals when deciding whether to donate to an NGO. Nonprofits fundraising internationally may therefore rely on framing to control perceptions [@ChaudhryHeiss:2021]. Below, we theorize the effect of different frames surrounding a country's legal environment and nonprofit organizational characteristics on individual donor preferences.

@@

One common signal for donors is organizational efficacy or the belief that an individual's donation will make a real difference. This can manifest itself through multiple mechanisms such as the amount an organization spends on overhead costs, whether an organization makes its finances public, or whether an organization participates in voluntary accountability practices, such as listing organizational details on websites like GuideStar and Charity Navigator.

An NGO's financial transparency, defined as the degree of completeness of information, provided by organizations to the public concerning its business activities, is one tangible heuristics that donors use to make their decisions. NGOs can engage in financial transparency in a variety of ways such as distribution of audited financial statements, third party intermediaries. Research shows that transparency is associated with greater donations, lower debt and better governance [@HarrisNeely:2021; @SaxtonKuoHo:2012]. It can also enhance donor trust and confidence in organizations [@Vaccaro:2006; @VaccaroMadsen:2006].

@@

While NGOs have control over organizational factors such as transparency and accountability, and can work to improve these characteristics and public perception of them, these NGOs may have little to no control over structural factors that can also influence philanthropy. These structural factors include political contexts in NGO host countries, especially their environments for civil society organizations. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations [@GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022; @Heiss:2017; @DupuyRonPrakash:2016]. As a result, NGOs must spend more time, effort, and resources to ensure their survival—at the expense of pursuing their missions.

Some NGOs have adapted to their repressive environments by recreating their organizations and changing their mission to avoid directly confronting the government. Analyzing the impact of anti-NGO laws in Bangladesh and Zambia, scholars have similarly found that NGOs with a broad mission ended up restricting themselves to service work and giving up on their advocacy work. Those that continued to engage in transnational advocacy changed the targets of their advocacy, the issues they focused on, and the language they used. For instance, many NGOs stopped lobbying the central government and focused their advocacy efforts on lower levels of government [@FransenDupuyHinfelaar:2021]. Similarly, in Russia, many NGOs working with foreign partners and lobbying the central government no longer consider it an effective strategy [@SundstromHenrySperling:2022]. Overall, fewer groups may end up focusing on contentious causes, instead focusing on regime-compatible causes.

The changes in civic environments impacting NGO activity have also influenced official donor responses. In response to the passage of these laws, scholars have found that while multilateral donors do not reduce aid following the passage of anti-NGO laws, they are associated with a 32% decline in bilateral aid inflows [@DupuyPrakash:2018]. Further, advocacy-oriented donors---i.e., those that fund democracy and civil society promotion activities---reduced their spending by 74%, but did not cut their spending on development projects such as education, health, water, and sanitation [@RightSpringmanWibbels:2022]. However, previous literature has established that individual private donors may not necessarily use the same metrics as official donors [@DesaiKharas:2018; @ChaudhryHeiss:2021]. Private foreign aid does not have to deal with the same strategic and political considerations as official aid and therefore, may be in a better position to respond to recipients' needs on the grounds [@EasterlyWilliamson:2011]. These anti-NGO laws may signal to private donors that NGOs undertake crucial work abroad, which is why governments perceive them as threatening and crack down on them. Therefore, we expect that government restrictions abroad will increase respondents' likelihood of donating to the organization.

@@

Identifying the causal link between structural factors and donor behavior is more difficult than measuring the link between organizational factors and donor behavior because (1) NGOs have less direct control over their political environments, and (2) organizational characteristics like financial transparency and accountability can sour the NGO-government relationship---governments may be more likely to target NGOs that lack financial transparency and accountability. Governments do not enact anti-NGO laws in isolation. Countries are more likely to repress NGOs when organizational features are in tension with government preferences and when organizations receive substantial funding from abroad [@DupuyRonPrakash:2015]. Therefore, the same organizational characteristic that can increase the likelihood of government crackdown can also simultaneously serve as a donor heuristic and influence perceptions of NGO deservingness.

@@

A conjoint research design also allows us to calculate treatment effects even if respondents do not see every combination of organization attributes, and correspondingly provides a substantial increase in statistical power. In standard factorial experimental designs, every combination of treatment attributes would need to be shown to some number of participants would ordinarily need to be shown all possible combinations of experimental attributes. In our experiment, we presented respondents with six possible treatments with varying counts of attributes for each treatment, yielding 576 possible unique combinations of features [EXPLAIN HERE THAT WE HAVE ALL THE IVS BUT WE LOOK AT JUST THESE]. Even with respondents answering twelve iterations of the experiment, not every combination was seen. However, as long as all the possible organizational attributes are well randomized and there are no systematic biases toward specific options (i.e. more respondents select the first option because it is the first) or toward earlier iterations of attribute choices (i.e. respondents are more careful and attentive for the first hypothetical organization than the last), we can pool all observations together for specific attributes of interest while marginalizing across all other attributes [@KertzerRenshonYarhi-Milo:2021]. This allows us to (1) estimate the effect of each experimental treatment even if some unique combinations of attributes were unseen, and (2) use a much smaller sample size than would be required in a more traditional factorial design.

@@

# Results

@@

## Discussion

@@

Over the last two decades, governments globally have systematically controlled and repressed NGOs using both violent tactics and administrative means over the last three decades---a phenomenon known as "closing" or "shrinking" civic space. This trend of closing civic space has important implications for local and international NGOs as well as donors. Recent research has analyzed the impact of this crackdown on public or official aid donors [@DupuyPrakash:2018; @Herrold:2020; @RightSpringmanWibbels:2022]. However, we know relatively little about how state repression of NGOs affects the preferences of foreign private donors [@ChaudhryHeiss:2021]. The responses of elite donors and foundations to government crackdown on civil society organizations remains a ripe area for research. Unlike state donors, individual and elite donors may not necessarily withdraw or reduce support for NGOs facing harassment abroad. This is important because private philanthropy towards organizations working in international affairs continues to grow despite inflation---from 2021 to 2022, individual giving to nonprofits in international affairs grew by over 10.9% in the US [@GivingUSA:2023].

Given the increasing hurdles faced by these nonprofits, how do individual donors in the U.S. feel about donating to legally besieged NGOs abroad? How do organizational-level factors such as financial transparency and accountability compare with more structural-level factors such as host country politics when deciding to donate? We argue that informational heuristics about NGOs’ legal and political environments and organizational-level attributes can influence donor preferences for international giving. Using a conjoint experiment with a nationally representative sample of the U.S. population, we find that both financial transparency and accountability increases individuals' willingness to donate NGOs. Learning about host government criticism and crackdown against NGOs decreases the likelihood of donation by itself; however, organizational features such as financial transparency and accountability protect against this dampening effect. INSERT LINE/MAKE MORE SPECIFIC.

These results have important implications for NGO operations, fundraising and survivability at a time when a number of INGOs are dealing with hostile host governments. A comparison of the effects of these organizational and structural factors is critical to NGO fundraising---while NGOs have control over transparency and accountability, as well as public perception of these organizational characteristics, they may little to no control over host government crackdowns over their organizations. As a result, they may need to rely entirely on improving individual donor perception of the causes of such crackdown, its impact on NGO operations and the need for private donor funds. This article provides an insight into the importance of different framing or informational heuristics that can motivate such individual donor giving.

Government crackdown on civil society organizations is even more challenging for local NGOs in Global South countries, where much government and public support is channeled via service delivery organizations. Local groups tackling contentious issues may not be able to raise funds domestically, as publics may prefer donating to service organizations focusing on health, education, and poverty [@AbsarCrowRon:2017]. Even local philanthropists may be deterred from giving to organizations focusing on contentious causes such as advocacy, media freedom, and anti-corruption initiatives due to poor tax incentives or a fear of retribution [@Brechenmacher:2017; @DupuyRonPrakash:2015; @Baoumi:2016]. Our results highlight the importance of organizational characteristics like transparency and accountability even in an era of closing civic space. By finding new ways to communicate about organizational transparency and accountability to potential donors, as well as the causes of government crackdowns on even transparent organizations, NGOs can help overcome individual donors' concerns about host governments' criticisms of and crackdown on these groups.

Future research should test how well these results map on to elite individual donors, not just the average private donor, as well as determine the extent to which these preferences translate into behavior. Future research should also test the generalizability of these findings to non-American populations. Donors in European countries may have different preferences when choosing to donate to NGOs, as nonprofits in the European sector typically do not rely as much on government funding, and already receive more support from private donors compared to the American context.
